---
title: "Exercises Data Science with R"
author: "Thomas Kirschstein"
output:
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    number_sections: false
---

# Chapter 1

There are quite a lot standard basic mathematical functions. Try to google for R cheat sheets if you are interested in a comprehensive overview.

```{r, exe1, echo=T, include=T, collapse=TRUE}
3^2     #1
sqrt(9) #2
9^.5
pi^2    #3
(abs(3^2-4^2))^.5 #4
log(exp(4), base = exp(1)) #5
log(exp(4))
log(100, base = 10) #6
log10(100) 
factorial(8) #7
exp(factorial(3)) #8
```

# Chapter 2

The chapter intends to focus on writing simple functions in R.  

1. Formulate the EOQ formula in R
```{r, exe2-1,  echo=T, include=T, collapse=TRUE}
cost_eoq_fun <- function(q, d, co, cl) {
  # returns total cost per period
  # d...demand
  # q...lot size
  # co...ordering cost
  # cl...stock holding cost
  ((1/2)*cl*q)+((d/q)*co)
}

# test cost function
cost_eoq_fun(d=100, q=20, cl = .1, co = 50)

eoq_fun <- function(co, d, cl) {
  # return optimal lot size
  # d...demand
  # co...ordering cost
  # cl...stock holding cost
  sqrt((2*co*d)/cl)
}
# optimal lot size
q.star <-  eoq_fun(d = 100, cl = .1, co = 50)
# optimal cost
cost_eoq_fun(d=100, q=q.star, cl = .1, co = 50)
```

2. Derive a function for calculating weighted Euclidean distance between two points.
```{r, exe2-2,  echo=T, include=T, collapse=TRUE}
weuc_d2_func <- function(x,y,w) {
  # calculates weighted Euclidean distance between x and y
  # y,x...vectors
  # w.. weight vector
  sqrt(sum(w*(x-y)^2))
}
# test distacne function
weuc_d2_func(x = c(1,2,3), y= c(3,2,1), w=c(1,1,1) )
# result should be sqrt(8)
```

3. Alter your EOQ function by checking whether all arguments are supplied and stop execution while displaying a error message.
-> postponed

4. Formulate a function for the Geometric Poisson density distribution.
```{r, exe2-3,  echo=T, include=T, collapse=TRUE}
geom_pois_dens_fun <- function(n, lambda, theta){
  # calculates density value of geometric Poisson distribution
  # n...integer, demand/successes, theta,lambda..parameters
  k.vec <- 1:n
  sum(exp(-lambda)*lambda^k.vec/factorial(k.vec)*(1-theta)^(n-k.vec)*choose(n-1, k.vec-1))
}
# test function
geom_pois_dens_fun(n=3, lambda=.5, theta = 2)

```


# Chapter 3

Basics on data types and data manipulation

1. Calculate the outer product of two vectors (without outer())
```{r, exe3-1,  echo=T, include=T, collapse=TRUE}
x <- 1:5
y <- 10:6
as.matrix(x) %*% t(as.matrix(y))
```


2. Define a function that calculates the trace of a matrix.
```{r, exe3-2,  echo=T, include=T, collapse=TRUE}
trace_func <- function(z){
  # calculates trace of z
  # z...matrix
  sum(diag(z))
} 
A <- matrix(rnorm(9) , ncol=3)
trace_func(A)
```


3.Create a vector containing the first 100 Fibonacci numbers.
Most commonly, the Fibonacci numbers are defined recursively by $F_n=F_{n-1}+F_{n-2}$ whereby $F_0 =0$ and $F_1=1$. However, there is also an explicit formulation: $F_n = \sum_{k=0}^{\lfloor \frac{n-1}{2} \rfloor} \binom{n-k-1}{k}$ (ceck [here](https://en.wikipedia.org/wiki/Fibonacci_number#Mathematics))

```{r, exe3-3,  echo=T, include=T, collapse=TRUE}
fib_num_fun <- function(n){
  # calculate nth Fibonacci number
  # n...number
  k.vec <- 0:floor((n-1)/2)
  sum(choose(n-k.vec-1, k.vec))
}
# vectorize fib_num_fun such that it accepts input vectors
vfib_num_fun <- Vectorize(fib_num_fun)
# doesn't work
fib_num_fun(1:100)
# works
vfib_num_fun(1:100)
```

4. Create a matrix containing the all binominal coefficients up to n = 50
```{r, exe3-4,  echo=T, include=T, collapse=TRUE}
pas <- outer(1:50, 1:50, choose)
```

5. Preference matrices in the Analystical Hirachy Process (AHP) show a form of (inverted) symmetry. How can you check this in R

The AHP method requires for a matrix $A=\{a_{ij}|i,j=1,...,n\}$ that $a_{i,j} = \frac{1}{a_{j,i}}$. Thereby, diagonal elements are always 1 (i.e., $a_{ii} = 1$)

```{r, exe3-5,  echo=T, include=T, collapse=TRUE}
# create 3x3 matrix 
A <- matrix(sample(1:10 , 9), ncol=3)
# diagonals to 1
diag(A) <- 1
# checks matrix elementwise
A == t(1/A) # -> problem double counting of elements
# Alternative: compare just elements of triangle sub-matrices
A[lower.tri(A)] == 1/A[upper.tri(A)]
```

6. Calculate the synthesis of an AHP preference matrix

The synthesis is calculated in two steps: 
(a) normalizing $A$ by dividing with column sums
(b) calculate row means of normalized matrix 

```{r, exe3-6,  echo=T, include=T, collapse=TRUE}
# convert A (chunk before) into a consistent matrix
A[lower.tri(A)] <- 1/A[upper.tri(A)]
# step (a) of synthesis
dev <- colSums(A) # calculate colSums
A.prime <- t(t(A) / dev) # Alternative a: using that R divides columnwise
A.prime <- A %*% diag(1 / dev) # Alternative b: use matrix calculation
# step (b) of synthesis
rowMeans(A.prime)
```

# Chapter 4

1. Construct a list with 100 entries. Afterwards, display only entries with odd indices.
```{r, exe4-1,  echo=T, include=T, collapse=TRUE}
x <- sample(x = 1:100, size = 100)
l1 <- as.list(x)
l1.odd <- l1[1:100 %% 2 == 1]
```


2. Construct a tibble or data frame consisting of 100 columns and 1 row. Extract only every 3rd column.
```{r, exe4-2,  echo=T, include=T, collapse=TRUE}
z <- matrix(sample(x = 1:1000, size = 100), nrow=1)
z <- as.data.frame(z)
z.sub <- z[, col(z) %% 3 == 0]
```

3. Construct a matrix of size $6 \times 6$ and fill it by sampling numbers between 1 and 100. Retain indices of all entries $\leq 50$.

```{r, exe4-3,  echo=T, include=T, collapse=TRUE}
A <- matrix(sample(x = 1:100, 36), ncol = 6)
which(A <= 50)
```

4. Formulate a function for calculating the moving average over a vector.

```{r, exe4-4,  echo=T, include=T, collapse=TRUE}
ma_fun <- function(x, n){
  # calculates moving average
  # x...vector to be averaged, n...time window width
  cx <- c(0,cumsum(x))
  (cx[(n+1):length(cx)] - cx[1:(length(cx) - n)]) / n
}
# test function
ma_fun(x = 1:10 , n=4)
```


# Chapter 5

1. Sample 100 observations in the range $[-3,3]$ from the follwing model and plot the sample:
$$ y = \sin(3 \cdot x^2 - 4) \cdot x + 0.5 \cdot x + \epsilon $$ whereby $\epsilon \sim N(0,0.5)$.
Can you estimate a regression function?

```{r, exe5-1,  echo=T, include=T, collapse=TRUE}
n <- 100                                # number of points to plot
x <- runif(n, min=-3, max=3)            # 100 x coordinates
y <- sin(3 * x^2 - 4) * x + 0.5 * x + rnorm(n, mean = 0, sd = sqrt(.5)) 
plot(x[order(x)], y[order(x)], xlab ="x", ylab = "y", pch = 16)
lin.reg <- lm(y~x)                      # fit a linear regression model
abline(lin.reg, col = "red", lwd =2)    # ... and plot the regression line
```

2. Plot the EOQ model in an appropriate range.

```{r, exe5-2,  echo=T, include=T, collapse=TRUE}
q.vec <- seq(265, 365, length.out = 100)
y.vec <- cost_eoq_fun(q = q.vec , d = 100, cl = .1, co = 50)
plot(q.vec, y.vec, xlab ="lot size q", ylab = "total cost", type = "l", lwd =2)
abline(v = eoq_fun(d = 100, cl = .1, co = 50), col="darkgrey", lwd=2, lty=2)
```

3. Plot the newsvendor model in an appropriate range.

The total cost in the normal newsvendor model is given by
$$ Z(q) = \left(c_u + c_o \right) \cdot \sigma \cdot \left( \varphi(q') + q' \cdot \Phi(q') \right) - c_u \cdot (q-\mu) $$ 
with $q' = \frac{q-\mu}{\sigma}$

```{r, exe5-3,  echo=T, include=T, collapse=TRUE}
newsven_obj_fun <- function(x, cu, co, mu, sigma){
      q.prime <- (x-mu)/sigma
	     (cu + co) * sigma * (dnorm(q.prime) + q.prime * pnorm(q.prime)) - cu * sigma * q.prime
}
q.vec <- seq(110, 125, length.out = 100)
cost.vec <- newsven_obj_fun(x = q.vec , mu = 100, cu = 2, co = 0.5, sigma = 20)
plot(q.vec, cost.vec, xlab ="lot size q", ylab = "total cost", type = "l", lwd = 2)

```

4. Plot the EOQ model in an appropriate range and add holding cost function and ordering cost function in different colors. Add a legend to the plot.

```{r, exe5-4,  echo=T, include=T, collapse=TRUE}
q.vec <- seq(265, 365, length.out = 100)
y.vec <- cost_eoq_fun(q = q.vec , d = 100, cl = .1, co = 50)
plot(q.vec, y.vec, xlab ="lot size q", ylab = "total cost", type = "l", lwd =2, ylim=c(12,max(y.vec)))
abline(v = eoq_fun(d = 100, cl = .1, co = 50), col="darkgrey", lwd=2, lty=2)
lines(x=q.vec, y=(1/2)*.1*q.vec, lwd=2, col="blue")
lines(x=q.vec, y=(100/q.vec)*50, lwd=2, col="red")
legend("left", lty = c(1,1,1,2), col = c("black","red","blue","darkgrey"), legend = c("total cost", "ordering cost", "storage cost", "optimal lot size"), bty="n" )
```


5. Consider the joint economic lot size model. Supplier and customer face  setup costs ($c^{sup.}_{set}$ ,$c^{cust.}_{set}$) as well as stock-holding cost rates ($c^{sup.}_{sh}$, $c^{sup.}_{sh}$). The demand rate of the customers is $\lambda$ and production rate of the supplier is $\mu$ (with $\lambda < \mu$). The cost functions of supplier and customer are
$$C^{cust.}(q) = \frac{\lambda}{q} \cdot c^{cust.}_{set}  + c^{cust.}_{sh} \cdot  \frac{q}{2} $$ and
$$C^{supp.}(q) = \frac{\lambda}{q} \cdot c^{sup.}_{set}  + c^{sup.}_{sh} \cdot  \frac{q}{2} \cdot \frac{\lambda}{\mu} . $$
Plot the cost functions of supplier, customer, and the SC (i.e., total cost function) in different colors. Add a legend to the plot.

```{r, exe5-5,  echo=T, include=T, collapse=TRUE}
jel_func <- function(q, c.supp.set, c.supp.sh, c.cust.set, c.cust.sh, lambda, mu){
  # costs in joint economic lot sizing model
  if(lambda > mu) stop("infeasible (lambda > mu)") # check feasibility
  c.cust <- lambda/q * c.cust.set  + c.cust.sh *  q/2 
  c.supp <-  lambda/q * c.supp.set  + c.supp.sh * q/2 * lambda/mu
  c.sc <- c.supp + c.cust
  return(data.frame(cost.customer = c.cust, cost.supplier = c.supp, cost.sc = c.sc))
}

q.vec <- seq(50, 200, length.out = 100)
y.df <- jel_func(q = q.vec, c.supp.set = 240, c.supp.sh = 4, c.cust.set = 100, c.cust.sh = 5, lambda = 100, mu = 160)

plot(q.vec, y.df$cost.sc, xlab ="lot size q", ylab = "costs", type = "l", lwd = 2, ylim=range(y.df))
lines(x = q.vec, y = y.df$cost.customer, lwd = 2, col = "blue")
lines(x = q.vec, y = y.df$cost.supplier, lwd = 2, col = "red")
abline(v = q.vec[which.min(y.df$cost.customer)], col = "blue", lwd = 1, lty = 2)
abline(v = q.vec[which.min(y.df$cost.supplier)], col = "red", lwd = 1, lty = 2)
abline(v = q.vec[which.min(y.df$cost.sc)], col = "darkgrey", lwd = 1, lty = 2)
legend("top", lty = c(1,1,1,2), col = c("black","red","blue","darkgrey"), legend = c("SC", "supplier", "customer", "opt. q"), bty="n" , horiz = T)
```

6. Consider the contract design problem with buy-back option (Thonemann, 2010, p. 479). Given are unit cost per product unit $c$ of the supplier, the sales price of the supplier $w$, the sales price of retailer $r$, the scrap revenue $v$ of the SC as well as normally distributed demand with  $y \sim N(\mu, \sigma^2)$. To be determined is the buy-back price $b$ such that the optimal profit of the SC is generated. The optimal SC profit is obtained if the retailer orders $q=\mu + \sigma \cdot \Phi^{-1}\left( CR \right)$ units, whereby $CR = \frac{r-c}{r-v}$ is the SC's critical ratio.  For the optimizing buy-back price $b$ holds: $$ b(w) =-r \cdot \frac{c-v}{r-c} + w \cdot \frac{r-v}{r-c}$$. The profits of retailer and supplier in case of optimal combinations of $b$ and $w$ holds $$G^{ret.}(w)=(r-w) \cdot \mu - (r-b) \cdot \sigma \cdot \Phi^{-1}\left( CR \right)$$ and $$G^{sup.}(w)=(w-c) \cdot \mu - (b-v) \cdot \sigma \cdot \Phi^{-1}\left( CR \right).$$
Plot the profit functions of retailer and supplier as well as the buy-back price function depending on $w$.

```{r, exe5-6,  echo=T, include=T, collapse=TRUE}
buy_back_func <- function(w, r, c, v, mu, sigma){
  # costs & buy-back price in buy-back model
  crit.rat.sc <- (r-c)/(r-v) # critical ratio SC
  z.sc <- qnorm(crit.rat.sc)
  b <- -(c-v)/(r-c)*r + (r-v)/(r-c)*w
  profit.cust <- (r-w)*mu - (r-b)*dnorm(z.sc)* sigma
  profit.supp <- (w-c)*mu - (b-v)*dnorm(z.sc)* sigma
  return(data.frame(buy.back.price = b, profit.supplier = profit.supp, profit.cust = profit.cust, profit.sc = profit.cust+profit.supp ))
}

w.vec <- seq(150, 250, length.out = 100)
y.df <- buy_back_func(w = w.vec, c = 100, r = 300, v = 10, mu = 100, sigma = 30)

plot(w.vec, y.df$profit.sc, xlab ="price w", ylab = "profit", type = "l", lwd = 2, ylim=range(c(y.df$profit.supplier,y.df$profit.cust,y.df$profit.sc)))
lines(x = w.vec, y = y.df$profit.cust, lwd = 2, col = "blue")
lines(x = w.vec, y = y.df$profit.supplier, lwd = 2, col = "red")
legend("bottomleft", col = c("black","red","blue"), lwd=2, legend = c("SC", "supplier", "customer"), bty="n")
```

7. Build a tidy data frame with `tibble()` containing two groups of observations: group A consisting of 100 observations with $x^A_i, y^A_i \sim \mathcal{N}(0,0.5)$ and group B consisting of 50 observations with $x^B_i, y^B_i \sim \mathcal{N}(3,1.5)$. Plot all points as a scatterplot and color the points groupwise.

8. The data set above is untidy. Create the same data set in a tidy way and recreate the plots above.

9. Alter the structured plot by connecting points by arrows. Test other smoothing methods offered by `geom_smooth`. Add three different smoothing functions to the plot.

10. Create a random sample $S = S_1 \cup S_2$ consisting of 50 obersations with $S_1 \sim \mathcal{E}(0.5)$ and 20 observations with $S_2 \sim  \mathcal{N}(3, 1)$. Plot a histogram of the sample overlaid with the theoretical exponential and normal distribution. (Hint: Use `rexp()` to sample from the exponential distribution.)

11. Plot the standard exponential distribution, mark 95%-quantile and highlight the area under the density curve beyond the 95%-quantile.

12. Try to recreate these violine plots from an untidy data set.

