---
title: "Exercises Data Science with R"
author: "Thomas Kirschstein"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Chapter 1

There are quite a lot standard basic mathematical functions. Try to google for R cheat sheets if you are interested in a comprehensive overview.

```{r, exe1, echo=T, include=T, collapse=TRUE}
3^2     #1
sqrt(9) #2
9^.5
pi^2    #3
(abs(3^2-4^2))^.5 #4
log(exp(4), base = exp(1)) #5
log(exp(4))
log(100, base = 10) #6
log10(100) 
factorial(8) #7
exp(factorial(3)) #8
```

# Chapter 2

The chapter intends to focus on writing simple functions in R.  

1. Formulate the EOQ formula in R
```{r, exe2-1,  echo=T, include=T, collapse=TRUE}
cost_eoq_fun <- function(q, d, co, cl) {
  # returns total cost per period
  # d...demand
  # q...lot size
  # co...ordering cost
  # cl...stock holding cost
  ((1/2)*cl*q)+((d/q)*co)
}

# test cost function
cost_eoq_fun(d=100, q=20, cl = .1, co = 50)

eoq_fun <- function(co, d, cl) {
  # return optimal lot size
  # d...demand
  # co...ordering cost
  # cl...stock holding cost
  sqrt((2*co*d)/cl)
}
# optimal lot size
q.star <-  eoq_fun(d = 100, cl = .1, co = 50)
# optimal cost
cost_eoq_fun(d=100, q=q.star, cl = .1, co = 50)
```

2. Derive a function for calculating weighted Euclidean distance between two points.
```{r, exe2-2,  echo=T, include=T, collapse=TRUE}
weuc_d2_func <- function(x,y,w) {
  # calculates weighted Euclidean distance between x and y
  # y,x...vectors
  # w.. weight vector
  sqrt(sum(w*(x-y)^2))
}
# test distacne function
weuc_d2_func(x = c(1,2,3), y= c(3,2,1), w=c(1,1,1) )
# result should be sqrt(8)
```

3. Alter your EOQ function by checking whether all arguments are supplied and stop execution while displaying a error message.
-> postponed

4. Formulate a function for the Geometric Poisson density distribution.
```{r, exe2-3,  echo=T, include=T, collapse=TRUE}
geom_pois_dens_fun <- function(n, lambda, theta){
  # calculates density value of geometric Poisson distribution
  # n...integer, demand/successes, theta,lambda..parameters
  k.vec <- 1:n
  sum(exp(-lambda)*lambda^k.vec/factorial(k.vec)*(1-theta)^(n-k.vec)*choose(n-1, k.vec-1))
}
# test function
geom_pois_dens_fun(n=3, lambda=.5, theta = 2)

```


# Chapter 3

Basics on data types and data manipulation

1. Calculate the outer product of two vectors (without outer())
```{r, exe3-1,  echo=T, include=T, collapse=TRUE}
x <- 1:5
y <- 10:6
as.matrix(x) %*% t(as.matrix(y))
```


2. Define a function that calculates the trace of a matrix.
```{r, exe3-2,  echo=T, include=T, collapse=TRUE}
trace_func <- function(z){
  # calculates trace of z
  # z...matrix
  sum(diag(z))
} 
A <- matrix(rnorm(9) , ncol=3)
trace_func(A)
```


3.Create a vector containing the first 100 Fibonacci numbers.
Most commonly, the Fibonacci numbers are defined recursively by $F_n=F_{n-1}+F_{n-2}$ whereby $F_0 =0$ and $F_1=1$. However, there is also an explicit formulation: $F_n = \sum_{k=0}^{\lfloor \frac{n-1}{2} \rfloor} \binom{n-k-1}{k}$ (ceck [here](https://en.wikipedia.org/wiki/Fibonacci_number#Mathematics))

```{r, exe3-3,  echo=T, include=T, collapse=TRUE}
fib_num_fun <- function(n){
  # calculate nth Fibonacci number
  # n...number
  k.vec <- 0:floor((n-1)/2)
  sum(choose(n-k.vec-1, k.vec))
}
# vectorize fib_num_fun such that it accepts input vectors
vfib_num_fun <- Vectorize(fib_num_fun)
# doesn't work
fib_num_fun(1:100)
# works
vfib_num_fun(1:100)
```

4. Create a matrix containing the all binominal coefficients up to n = 50
```{r, exe3-4,  echo=T, include=T, collapse=TRUE}
pas <- outer(1:50, 1:50, choose)
```

5. Preference matrices in the Analystical Hirachy Process (AHP) show a form of (inverted) symmetry. How can you check this in R

The AHP method requires for a matrix $A=\{a_{ij}|i,j=1,...,n\}$ that $a_{i,j} = \frac{1}{a_{j,i}}$. Thereby, diagonal elements are always 1 (i.e., $a_{ii} = 1$)

```{r, exe3-5,  echo=T, include=T, collapse=TRUE}
# create 3x3 matrix 
A <- matrix(sample(1:10 , 9), ncol=3)
# diagonals to 1
diag(A) <- 1
# checks matrix elementwise
A == t(1/A) # -> problem double counting of elements
# Alternative: compare just elements of triangle sub-matrices
A[lower.tri(A)] == 1/A[upper.tri(A)]
```

6. Calculate the synthesis of an AHP preference matrix

The synthesis is calculated in two steps: 
(a) normalizing $A$ by dividing with column sums
(b) calculate row means of normalized matrix 

```{r, exe3-6,  echo=T, include=T, collapse=TRUE}
# convert A (chunk before) into a consistent matrix
A[lower.tri(A)] <- 1/A[upper.tri(A)]
# step (a) of synthesis
dev <- colSums(A) # calculate colSums
A.prime <- t(t(A) / dev) # Alternative a: using that R divides columnwise
A.prime <- A %*% diag(1 / dev) # Alternative b: use matrix calculation
# step (b) of synthesis
rowMeans(A.prime)
```

# Chapter 4


1. Construct a list with 100 entries. Afterwards, display only entries with odd indices.
```{r, exe4-1,  echo=T, include=T, collapse=TRUE}
x <- sample(x = 1:100, size = 100)
l1 <- as.list(x)
l1.odd <- l1[1:100 %% 2 == 1]
```


2. Construct a tibble or data frame consisting of 100 columns and 1 row. Extract only every 3rd column.
```{r, exe4-2,  echo=T, include=T, collapse=TRUE}
z <- matrix(sample(x = 1:1000, size = 100), nrow=1)
z <- as.data.frame(z)
z.sub <- z[, col(z) %% 3 == 0]
```

3. Construct a matrix of size $6 \times 6$ and fill it by sampling numbers between 1 and 100. Retain indices of all entries ≤ 50.

```{r, exe4-3,  echo=T, include=T, collapse=TRUE}
A <- matrix(sample(x = 1:100, 36), ncol = 6)
which(A <= 50)
```

4.Formulate a function for calculating the moving average over a vector.

```{r, exe4-4,  echo=T, include=T, collapse=TRUE}
ma_fun <- function(x, n){
  # calculates moving average
  # x...vector to be averaged, n...time window width
  cx <- c(0,cumsum(x))
  (cx[(n+1):length(cx)] - cx[1:(length(cx) - n)]) / n
}
# test function
ma_fun(x = 1:10 , n=4)
```






