---
title: "Exercises project"
author: "Thomas Kirschstein"
output:
  html_document:
    df_print: paged
---

# Data description 

The data set "biathlon_results_women.csv" contains results of all womens' biathlon races of the IBU World Cup, the World Championships, and at the Olympic Games from season 2008/09 till 2017/18.  The raw data is extracted from [IBU results homepage](https://biathlonresults.com/) where also a description of the variables is provided.

The data set contains 25 variables as follows

name | description
------------- | -------------
date	| race date in d.m.year format
season	| season id
competition	| Type of competition (I...Individual, S...Sprint, M...Mass start, P...Pursuit)
type	| Type of race (W...world Cup, O...Olympic, C...World Championship)
name	| Name of athlete
nation	| nation of athlete
rank	| final rank (total time)
course.rank	| final rank (course time)
tot.time	| total time at finish (in seconds)
course.time	| course time at finish (in seconds)
location	| race location
elevation	| elevation of location (in m)
height.diff	| height difference of course profile (in m)
max.climb	| maximum climb in course profile (in m)
total.climb	| total climb over course (in m)
birthday	| birthday of athlete
snow.conditions	| snow conditions (check IBU website, [wiki](https://en.wikipedia.org/wiki/Classifications_of_snow), [blog](http://blog.tahoemountainsports.com/tahoe-cross-country-skiing-well-its-about-time/))
snow.temperature	| snow temperature (in $^\circ$C)
air.temperature	| air temperature (in $^\circ$C)
weather.conditions	| weather conditions (self-explaining)
fails.total	| total number of missed targets in shooting
speed	| skiing speed, i.e. course length/course time (in m/s)
age	| age of athlete at race time (in years)
home	| whether it was a home race

# Task

Compose a report with a descriptive analysis of the data. Investigate which variables affect the skiing speed of the athletes. Use basic statistics as well as appropriate plots to illustrate the effects of environmental conditions or course profiles on skiing speed. 

# Read data 

At first read the data set into the workspace either via the console or the "import dataset" dialog in environment menu (right upper pane). 

```{r, proj-1, echo=TRUE, warning=TRUE, collapse=TRUE}
biathlon_results_women <- read.csv("~/Thomas/Owncloud/Data_Science_PL/R-Skripte/DatsciWiR_exercise/exercise_case_study/biathlon_results_women.csv")
```

The data should look like this:

```{r, proj-2, echo=TRUE, warning=TRUE, collapse=TRUE}
library(DT)
datatable(head(biathlon_results_women, 50))
```

